\section{The allocation-channel study, in brief}
\label{app:companion}

% The allocation-channel study is OUR OWN prior experiments, reported
% first-hand here — no external reference, per the author's decision
% 2026-07-09. If that study is ever submitted as its own paper, disclose
% the overlap to both venues.
% src: figures/summary.md a.k.a. blindness_results.json (Phase 1);
%      h2_unified/report/results.json (voids, 5 paired seeds);
%      crossover/report/results.json (torus incl. B5).
% MOVED to the supplementary (3DV page limit, 2026-07-13) exactly as the
% NOTE(author) here anticipated; \appendix now lives in supplementary.tex.
% Anonymity: "our" is double-blind-safe — the study is unpublished and
% travels with this submission; it names no authors.

This appendix reports, with their numbers, the three findings of our
allocation-channel study --- controlled resampling-prior experiments on the
same pipeline, scenes, and budgets, conducted before moving topology into
the objective --- that this paper builds on.

\paragraph{(1) Geometric metrics are topology-blind.} Three constructed
cases each pit a topologically correct candidate A against a wrong one B
whose geometry is bisection-matched, making Chamfer \emph{equal by
construction} (40k samples, metrics normalized by the ground-truth bounding
diagonal; Table~\ref{tab:blindness}). The discriminating-dimension
bottleneck separates every pair --- by $35$--$40\times$ in the H0/H2 cases
--- while in the first two cases Hausdorff$_{95}$ actually \emph{prefers}
the topologically wrong candidate. Chamfer cannot tell a benign bump from a
destroyed void; the persistence metric can.
% src: figures/summary.md (Phase-1 result table, 3/3 gate)

\begin{table*}[t]
\centering
\caption{Blindness probes: A topologically correct, B wrong, geometry
bisection-matched so Chamfer is equal by construction. Bottleneck to ground
truth in the discriminating dimension.}
\label{tab:blindness}
\small
\begin{tabular}{lcccc}
\toprule
case (disc.\ dim) & Cham.\% A $=$ B & bott.\ A & bott.\ B & ratio \\
\midrule
% src: figures/summary.md (Phase-1 result table)
thin handle, genus $0{\to}1$ (H1) & 0.916 & ${\approx}0$ & .0302 & --- \\
bridge merges two components (H0) & 0.628 & .0014 & .0574 & $40\times$ \\
punctured enclosed void (H2)      & 0.977 & .0040 & .1385 & $35\times$ \\
\bottomrule
\end{tabular}
\end{table*}

\paragraph{(2) The prior arms, and why voids are ``width-primary''.} The
study biases DiffSoup's own resampler with a precomputed importance field
built from the target's significant features. Each significant feature $f$
(lifetime $p_f$ above the significance threshold) contributes Gaussian
kernels at the coordinates $C_f$ of its birth/death-simplex vertices ---
the loop-filling triangle, void-filling tetrahedron, or component-merging
edge --- sized by the feature's own death scale:
\begin{equation}
\begin{gathered}
\phi_d(q)=\!\!\sum_{f:\,\dim f=d}\,\sum_{c\in C_f}\! w_f\,
  e^{-\lVert q-c\rVert^2/(2\sigma_f^2)},\qquad w_f = p_f/s^2,\\
\sigma_f = s\,\mathrm{clip}\!\big(\sqrt{\smash[b]{\mathrm{death}_f}},\,
  3r_{\mathrm{med}},\,\tfrac{1}{2}\mathrm{diag}\big),
\end{gathered}
\end{equation}
evaluated on a dense target-surface sample, normalized per dimension to
$[0,1]$ (99.5th percentile), combined by max; a query takes its nearest
surface sample's value. The spread knob $s$ widens each kernel while
preserving its injected surface mass ($\sigma\,{\to}\,s\sigma$,
$w\,{\to}\,w/s^2$, so $w\sigma^2$ is $s$-invariant): $s{=}1$ is the
concentrated prior (B2), $s{=}3$ the spread prior (B4) --- the field C5
reuses. The field enters resampling at strict budget parity through two
levers: prune protection --- faces drop in order of $\mathrm{vis} +
\lambda\,Q_{0.9}(\mathrm{vis})\cdot\mathrm{imp}$, removing exactly the
required count --- and respawn bias --- top-importance visible faces
subdivide first, then the protective order re-prunes to the cap. Remaining
arms: B0 baseline; B1 = B2 applied at initialization only; B3/B5
random-center non-topological controls width-matched to B2/B4 respectively.
% src: methods/topo_importance.py (build_importance_field, _localize_features)
% + methods/topo_resampling.py (keep_map, densify_and_rebalance)
On the void class
(Table~\ref{tab:companion}; $N{=}1200$, five paired seeds) B4 cuts the
bottleneck $33$--$53\%$ below baseline --- but the width-matched random
control B5 recovers most of that gain, leaving only a small, shape-dependent
topological residual (cylinder $4.6\sigma$, sphere $2.2\sigma$, cube tie);
one-shot placement washes out entirely (B1 $\approx$ B0, $.0546$ vs $.0535$
on the sphere). Hence the verdict quoted in the main paper's introduction:
the prior's topological value is carried largely by its spatial
\emph{width}.
% src: h2_unified/report/results.json (all five arms, paired seeds)

\begin{table}[t]
\centering
\caption{Allocation study, void class (H2), $N{=}1200$, five paired seeds:
tail bottleneck to target (lower is better). Spread beats concentrated;
the width-matched random control (B5) recovers most of the spread gain.}
\label{tab:companion}
\footnotesize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{lccccc}
\toprule
 & & B2 & B3 & B4 & B5 \\
shape & B0 & conc. & narrow rand. & spread & wide rand. \\
\midrule
% src: h2_unified/report/results.json (bott_tail_mean)
sphere   & .0535 & .0438 & .0440 & \textbf{.0357} & .0374 \\
cube     & .0579 & .0409 & .0403 & .0273 & \textbf{.0270} \\
cylinder & .0545 & .0440 & .0477 & \textbf{.0360} & .0402 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{(3) No prior shape repairs loops.} On the torus ($N{=}700$, five
paired seeds, true $\beta_1{=}2$) every arm fails the topology-specificity
test: the best condition is the \emph{non-topological} wide control B5
($.0267$), ahead of spread-topological B4 ($.0316$), baseline ($.0409$), and
narrow random B3 ($.0397$); the concentrated topological field B2 is worst
($.0425$) \emph{and} manufactures phantom handles --- final significant-H1
count $4.4$ against the true $2.0$, with the worst Chamfer ($1.32$ vs
baseline $1.01$). Spreading merely stops the harm; it never beats the random
control. That failure is the opening premise of the main paper, and the
class the loss channel wins there (main paper \S5.1, Table~1).
% src: crossover/report/results.json (torus B0/B2/B3/B4/B5 + final_nsig)
