\section{Mechanism and discussion}
\label{sec:discussion}

% TODO(human): voice pass; consider merging with results at venue length.

\paragraph{Sparse tugs vs.\ budget allocation.} Each significant pair
back-propagates through the $3$--$7$ sample positions of its birth and
death simplices, so the loss delivers \emph{precise but sparse} pulls,
iterated by re-pairing --- the toy probes show surgery ``walking'' around a
defect a few points at a time; the prior, by contrast, biases every
respawn decision diffusely across a feature's whole support. The
resampling prior is the opposite: broad, indirect, budget-limited. The
results sort cleanly along this axis: 1-D loops, which the prior could only
over- or under-tessellate, respond to precise value pressure (H1 win); wide
2-D shells benefit from both (H2: prior helps, loss helps more, composition
best); 0-D component structure needs neither --- any spreading pressure
suffices (H0 null, three channels).

\paragraph{Why the control damages topology.} The norm-matched repulsion
imposes a length-scale floor on the sampled surface; the void's death time
pins at that equilibrium (identical across seeds) and thin structures
collapse ($\beta_2$ loss on the cube, $\beta_1$ loss on the torus).
Fig.~\ref{fig:pdtraj} makes this visible: once the repulsed cloud reaches
its equilibrium spacing, no refinement can tighten the void closure --- the
death coordinate is pinned, then dragged, while near-diagonal phantom bars
accumulate. The loss has no such length-scale constraint: it pulls the
death coordinate itself and the geometry follows, rather than the other
way around. This sharpens the control argument: spending the same gradient
budget through the same channel \emph{without} topological information is
not neutral --- it is harmful --- so C1's wins cannot be an artifact of
extra regularization.

\paragraph{Measurability at the loss's density.} A persistence loss can only
enforce features that clear the significance floor at its sampling density
$M$: the torus void and the double torus's two tube loops do not, at
$M{=}2048$. We treat this as a principle rather than a nuisance: the target
bundle is built (and audited) at the same $M$, features below the floor are
excluded from the objective, and the double-torus stretch case is evaluated
against its two measurable loops --- which is also all the reconstruction
itself can express at feasible budgets. Raising $M$ buys resolution at
linear-ish persistence cost; the trade is explicit and logged at bundle-build
time.

The floor is quantifiable. A bar is significant iff its lifetime exceeds
$6\,r_{\mathrm{med}}(M)$ ($r_{\mathrm{med}}$ = median nearest-neighbour
spacing, the implementation's self-calibrating threshold), and
$r_{\mathrm{med}}$ follows the uniform-sampling law $\propto M^{-1/2}$
(measured $.0090{\to}.0028$ over $M{=}2048{\to}20000$; ratio $3.16$ vs
$\sqrt{9.77}{=}3.13$). This puts numbers on both stories in this paper. The
double torus's tube loops live at lifetime $.045$--$.046$ on the clean
surface: at $M{=}2048$ that is $0.83$--$0.85\times$ the floor --- invisible
by a $15\%$ margin, clearing it only from $M\!\approx\!3{\times}10^{3}$ ---
while at the evaluation density ($M{=}20000$, floor $.0171$) they clear it
$3.1\times$; the binding constraint there is the \emph{representation}
(the soup never expresses the tubes at any feasible $N$), so raising $M$
alone cannot close the gap. The torus quantifies the interaction with
reconstruction noise: its second loop clears the clean-surface floor by
only $1.33\times$, a margin the live cloud's noise consumes --- which is
exactly why that loop is chronically sub-threshold during training and
recruitment becomes its only gradient path (the C6 ablation).
Appendix~\ref{app:diagnostics} tabulates the counts across $M$; at
$M{=}4096$ all four double-torus loops clear the floor. Adaptive or
multi-scale density --- raising $M$ where features sit near the floor, or
hierarchical floors --- is the natural extension; the trade is explicit,
since persistence cost grows $O(M\log M)$.
% src: experiments/density_bound.py — persistence_from_target on the GT
% meshes at M in {2048, 20000}, seed 0 (2026-07-09); M-sweep in App. B

\paragraph{Recruitment is landscape-guided in practice --- and it is the
mechanism for loops.} Its location blindness --- pull the nearest bar
anywhere --- did not misfire on our geometry because unproductive tears
self-extinguish (their bar's death saturates at the local detour scale)
while productive cuts run away toward the target. The caveat stands for
target features larger than any achievable detour; the probe and its
mechanism are reported so the term is used with eyes open. The C6 ablation
sharpens its role beyond cold-start repair: a loop that never clears the
live significance floor at $M$ receives metric pressure \emph{only} through
recruitment, at every refresh --- matching alone reduces the torus to
baseline. Recruitment is how the loss reaches below its own significance
floor on the live side.

\paragraph{Prescription.} Correct topology in the \emph{loss}; allocate
budget with a \emph{wide} prior; combine for the best of both. Curriculum
schedules are unnecessary once the loss scale is calibrated against the
photometric gradient.
