\section{Introduction}
\label{sec:intro}

% Framing: measurement -> paradigm test (suppl. §A) -> objective.
% TODO(human): voice pass.
% 3DV pass 2026-07-13: appendix refs -> supplementary; light trim only.
% Advisor round 3 (2026-07-16): allocation-vs-objective framing made
% explicit; research hypotheses RH1-3 named (RH labels avoid colliding
% with the homology classes H0/H1/H2); numbers + % src unchanged.

Differentiable rasterization reconstructs a triangle soup from images by
following photometric gradients~\citep{tojo2026diffsoup}, periodically
resampling under a fixed budget. Photometric losses are
topologically blind: reconstructions that differ in genus, connectivity, or
enclosed-void structure can be indistinguishable to Chamfer and Hausdorff
while the bottleneck distance between persistence diagrams separates them
$35$--$40\times$ on probes whose Chamfer is equal \emph{by construction}
(\S\ref{sec:related}; suppl.\ Table~S1). Fig.~\ref{fig:matrix} previews
the consequence: identical budgets, equal-or-better Chamfer, topology
standing or falling with the objective.
% src: figures/summary.md (Phase 1); h2_unified + crossover reports (suppl. §A)

% Round-5 (2026-07-19, advisor Related-Works doc): the 2025--26
% triangle-turn context woven in as a clause (graphics-pipeline
% compatibility) per the requested Introduction paragraph — adapted, not
% pasted; cites long-verified. The requested "our work addresses this
% gap" echo is ¶3's existing closing sentence (unchanged).
A decade of connectivity-free reconstruction --- point splats,
Gaussians, and lately a return to triangle soups for their native
graphics-pipeline
fit~\citep{held2025trianglesplatting,burgdorfer2025radiant,tojo2026diffsoup}
--- has treated topology errors as symptoms of geometric
under-provision: represent the surface better, or allocate the budget
more wisely, and topology should follow. The allocation half is
directly testable, and we tested it first, in a controlled study of
where the resampler spends triangles (suppl.\ \S A).
It fails: a topological prior helps enclosed voids, but any equally-wide
\emph{non-topological} prior reproduces most of the gain, and no prior
shape repairs loops --- concentrating manufactures phantom handles,
spreading never beats a random control.
% src: PHASE2B_CROSSOVER.md; h2_unified/report/results.json

The null result relocates the root cause: not \emph{where} triangles are
spent, but \emph{what the objective measures} --- a photometric loss is
a local, pixel-space signal; a loop or an enclosed void is a global,
categorical property of the surface. This paper therefore moves topology
into the \emph{objective}: a differentiable persistence term whose
gradients act directly on features' birth/death values rather than on
where triangles respawn.

% Round-4 restructure: the review asked for the RHs as a visually
% distinct numbered list; paid for by running the contribution bullets
% into a prose paragraph. Content unchanged.
Three research hypotheses structure the study, each tested by a
pre-designed condition:
\begin{enumerate}
\setlength{\itemsep}{1pt}
\item \textbf{RH1 (diagnosis)}: topology errors originate in the
  training objective, not in primitive allocation --- tested by the
  allocation study above.
\item \textbf{RH2 (specificity)}: a persistence term improves topology
  in ways no norm-matched non-topological regularizer through the
  identical channel can reproduce --- the control C2
  (\S\ref{sec:conditions}).
\item \textbf{RH3 (complementarity)}: objective and allocation are
  complementary channels, not substitutes --- the composition C5.
\end{enumerate}
Concretely, we contribute \textbf{a matched persistence loss with a
pair-frozen backward} (\S\ref{sec:method}): exact persistence forward
(GUDHI alpha complex), backward through closed-form circumradii ---
Gabriel-exact for 100\% of significant pairs in practice;
% src: PHASE3_PLAN.md Appendix A (22/22 gate)
\textbf{a recruitment term for missing features}: optimal matching
yields zero gradient when a target feature has no live counterpart;
recruitment attaches the nearest live bar, and removing it collapses
the loop-class win to baseline (\S\ref{sec:results}); \textbf{one calibrated knob}: $\lambda$ set by a
gradient-norm ratio $\rho$, flat over a $10\times$ range, no curriculum
needed --- an ablation, not an assumption;
% src: PHASE3_PLAN.md Appendix C (rho sweep), Appendix D (C3)
and \textbf{channel-controlled evaluation}: every claim tested against a
norm-matched \emph{non-topological} control through the identical
channel (a repulsion on the same surface samples) --- the allocation
study's discipline, inherited.

% Round-4 trim: multipliers live in the abstract + §5; the intro preview
% keeps only the two claims unique to it (RH2 loop win; RH1 sharpened by
% the H0 null).
The loss proves topology-specific on the class no allocation prior ever
won --- loops, with zero phantom handles --- confirming RH2; the two
channels compose (RH3); and the H0 component class is matched by the
generic control, sharpening RH1: topological guidance earns its keep
where geometry and topology decouple.
% src: PHASE3_PLAN.md Appendix D / topo3/report/results.json
