\section{Limitations}
\label{sec:limitations}

% TODO(human): trim to venue length; keep the honesty.

\textbf{Synthetic, closed, single-machine.} All quantitative claims are on
synthetic renders of closed surfaces with known topology, on one GPU. The
scene family is analytic plus three external genus-known meshes
(\S\ref{sec:generality}) --- still short of a public shape-set benchmark ---
and a real (open-surface, clinical) demonstration remains future work; open
surfaces add a genuine research obstacle --- scan boundaries contaminate H1,
and no clean ground-truth diagram exists. Concrete paths exist: a
boundary-aware persistence definition (relative homology), filtering
boundary-born bars before matching, or template-aligned initialization
that suppresses boundary noise; we judge bar-filtering the pragmatic
near-term step, with a synthetic open-surface stress test (a ray-cast
solid plus point noise) as the natural first experiment.

\textbf{Density floor.} The loss cannot see features below the significance
threshold at its sampling density (torus void; two of four double-torus
loops at $M{=}2048$). The bundle audit makes this visible, not invisible ---
but it is a real ceiling on what the objective can enforce.

\textbf{Fixed target diagram.} The target bundle presumes known ground-truth
topology (true in reconstruction-from-scans settings where a prior scan or
template exists; not in blind capture). Three template-free designs are
natural but untested here: minimizing total persistence (risks collapsing
legitimate structure), enforcing a target \emph{genus} via a penalty on
the significant-H1 count (the most robust of the three, and the one we
would try first for blind capture), or statistical priors over a shape
class's diagrams.

\textbf{Control strength.} The repulsion control is aggressive (it destroys
topology at the calibrated magnitude); a gentler variant at half the radius
is still $1.6\times$ worse than baseline (sphere: $.0972$ vs $.0623$), so the
verdicts do not hinge on control tuning --- but control \emph{design} for
loss-channel studies deserves more variety than one family at two strengths.
% src: topo3/report/results.json

\textbf{Nondeterminism.} DiffSoup's CUDA training is not bit-reproducible:
divergence enters at resampling, where borderline prune/split decisions
flip under floating-point drift, so even baseline re-runs differ in
detail. Quantitatively, tail coefficients of variation run $7$--$10\%$ for
baseline and loss arms alike, and a loss-identical re-run pair (the sphere
C6 arm, which is C1 by construction there) landed $6\%$ apart --- the
scale of pure run-to-run noise. All statements are therefore seed-averaged
with Welch statistics.
% src: topo3/report/results.json (sd/mean per arm); C6-vs-C1 sphere
% ("Deferred arms" removed 2026-07-09 per advisor revision item 5.)

\textbf{Recruitment guarantees.} Recruitment's self-correction premise ---
unproductive edits saturate at the local detour scale --- is validated
empirically (toys, training, stress test), not proven; pathological
geometries (very thin filaments, tightly constrained manifolds) could
defeat it. A failure-case study and a hardness analysis of
recruitment-based matching are future work.
