% \mbox: keep "triangle" unhyphenated in the heading (it was breaking as
% "tri-angle" in the two-column measure); hyperref strips the \mbox in
% the PDF bookmark (warning class already filtered by cvpr.sty).
\section{Method: a matched persistence loss for \mbox{triangle} soups}
\label{sec:method}

% Grounded in methods/topological_loss.py + PHASE3_PLAN.md §3-4.
% TODO(human): tighten notation to venue conventions.
% 3DV pass 2026-07-13: display equations reflowed for the two-column
% measure (gathered/split); C5's appendix ref -> supplementary; wording
% tightened. All equations and numbers unchanged.

\subsection{Objective}
DiffSoup trains with a photometric objective $L_{\mathrm{photo}}$ (opacity
auxiliary + $0.8\,L_1 + 0.2\,\mathrm{SSIM}$)~\citep{tojo2026diffsoup}, which
we adopt \emph{unchanged}: every ablation below isolates the topology
channel, never photometric tuning. We add
\begin{equation}
L \;=\; L_{\mathrm{photo}} \;+\; \lambda(t)\, L_{\mathrm{topo}}(V),
\end{equation}
where $V$ are the soup's vertex positions and $L_{\mathrm{topo}}$ compares the
persistence diagram of the live surface to a fixed \emph{target bundle}
$\mathcal{T}$: the significant finite bars of the ground-truth shape's
diagram, computed once at the sampling density $M$ the loss will see
(alpha values shift with spacing: density matching is a correctness
requirement), normalized by the target's bounding-box diagonal, and
locked for all live evaluations.
% Round-4: "significant" was used ~30x before its §6 definition — the
% one-line definition now sits at first use; §6 keeps the scaling law.
Throughout, a bar is \emph{significant} iff its lifetime exceeds the
measurement floor $6\,r_{\mathrm{med}}(M)$ ($r_{\mathrm{med}}$ = median
nearest-neighbour sample spacing; derived and validated in
\S\ref{sec:discussion}).

\subsection{Live diagram and pair-frozen backward}
At each refresh we draw $M$ surface samples ($M{=}2048$ unless the
measurement-floor rule of \S\ref{sec:setup} dictates a denser bundle)
$X_j=\sum_i w_{ij}\,V[F(f_j),i]$ with faces $f_j$ drawn
$\sigma(\alpha)\times$area-weighted and barycentric weights $w_{ij}$ frozen;
$X$ is recomputed from live $V$ every step, so gradients reach vertices while
the combinatorics stay fixed. The alpha complex of
$X$~\citep{edelsbrunner1994alphashapes}, computed exactly with
GUDHI~\citep{maria2014gudhi}, yields persistence
pairs (birth simplex, death simplex) per dimension. For a Gabriel simplex the
alpha filtration value is its squared circumradius, so each paired bar's
coordinates are closed-form functions of a few sample positions, all
exactly differentiable:
\begin{equation}
\begin{gathered}
R_{\mathrm{edge}}=\tfrac12\lVert p_1{-}p_0\rVert,\qquad
R_{\mathrm{tri}}=\frac{\lVert u\rVert\,\lVert v\rVert\,\lVert u{-}v\rVert}
                      {2\,\lVert u\times v\rVert},\\[2pt]
R_{\mathrm{tet}}=\lVert y\rVert\quad\text{with}\quad
A\,y=\tfrac12\operatorname{diag}\!\big(AA^{\!\top}\big),
\end{gathered}
\end{equation}
where $u=p_1{-}p_0$, $v=p_2{-}p_0$, and $A$ stacks the rows $p_i{-}p_0$
(the circumcenter is $p_0{+}y$). Edges carry H0 deaths and H1 births,
triangles H1 deaths and H2 births, tetrahedra H2 deaths.
% src: methods/topological_loss.py circumradius_{edge,triangle,tetra}
We verify, per pairing, that the re-expressed value matches the filtration
value (relative $10^{-6}$); a pair failing this Gabriel test is
\emph{detached} --- it keeps its (correct) value in the loss but propagates
no gradient through that simplex, so an exactness failure can never inject
a wrong gradient.
In practice the gate never fires: 100\% of significant pairs are
Gabriel-exact on all six shapes' clean clouds, and the logged
per-refresh failure count is zero at every one of the $5{,}500{+}$
logged refreshes (including all $2{,}400$ noisy ones,
\S\ref{sec:results}) --- a constant $100\%$ pass rate, not an
aggregate. Stronger fallbacks (dropping pairs, subgradients, locally
raising $M$) remain available.
% src: PHASE3_PLAN.md Appendix A (mapping + Gabriel gates, 22/22);
% per-refresh sum recomputed 2026-07-17 over every topo_loss_log.json
% with the refresh schema: 5,560 refreshes, n_gabriel_fail = 0 throughout
% (scratchpad gabriel_scan.py over output/synth/**/topo_loss_log.json)
The pairing is treated as locally constant --- the standard pair-frozen
device~\citep{gabrielsson2020topologylayer,carriere2021optimizing} ---
refreshed every $K{=}10$ steps and, necessarily, after every resampling
event (which replaces the vertex and face tensors wholesale); $K{=}10$
is an amortization choice, not a tuned knob --- it bounds refresh cost
($174$\,ms each; the fixed ${\approx}46$\,s of \S\ref{sec:results})
while the pairing stays local.
% src: PHASE3_PLAN.md Appendix B-D (174 ms refresh; +46 s fixed cost)

\subsection{Matching, diagonal pressure, and recruitment}
Let $D$ be the live significant bars and $\mathcal{T}$ the target bars of one
dimension. An optimal partial matching (Hungarian on the standard augmented
cost; equivalently $W_2$) splits bars into matched pairs, unmatched live, and
unmatched target:
\begin{equation}\label{eq:matching}
\begin{split}
L_{\mathrm{topo}} = \!\!\sum_{(i,j)\,\mathrm{matched}}\!\!\!
  \big[(b_i-b^*_j)^2+(d_i-d^*_j)^2\big]\\[-2pt]
 \;+\; \sum_{i\,\mathrm{unmatched}} \Big(\tfrac{d_i-b_i}{2}\Big)^{\!2}
 \;+\; L_{\mathrm{recruit}} .
\end{split}
\end{equation}
The first term pulls matched features to their target coordinates; the second
pushes spurious features to the diagonal. The third exists because optimal
matching has a blind spot that is provable, not merely observed:
% Round-4: stated as a lemma (the review asked the provable fact to be
% separated cleanly from recruitment's *empirical* self-correction, §7).
\begin{lemma}\label{lem:zerograd}
If a target bar $t\in\mathcal{T}$ is unmatched under the optimal partial
matching, then $L_{\mathrm{topo}}$ without $L_{\mathrm{recruit}}$ is
locally independent of $t$: no component of its gradient in $V$ acts to
create the missing feature.
\end{lemma}
\noindent\emph{Proof sketch.} The unmatched target's augmented cost is its
diagonal penalty $\big((d^*_t-b^*_t)/2\big)^2$, constant in the live
coordinates; every $V$-dependent term of Eq.~\ref{eq:matching} references
live bars only, and the matching is locally constant in $V$ (the
pair-frozen regime) --- no gradient term is directed at $t$.\hfill$\square$

Recruitment restores a descent direction. Let $\mathcal{T}_{\mathrm{miss}}$ be the unmatched target
bars, processed in decreasing persistence order, and let the candidate pool
$\mathcal{A}$ hold every live bar not claimed by the matching --- unmatched
significant bars and, crucially, \emph{sub-threshold} bars, which matching
never sees. Each missing target greedily claims its nearest pool bar,
without replacement, under the same squared birth--death metric as the
matched term:
\begin{equation}
\begin{split}
L_{\mathrm{recruit}} &= \!\!\sum_{j\in\mathcal{T}_{\mathrm{miss}}}\!\!
  \big[(b_{k(j)}-b^*_j)^2+(d_{k(j)}-d^*_j)^2\big],\\
k(j) &= \operatorname*{arg\,min}_{k\,\in\,\mathcal{A}_j}
  \big[(b_k-b^*_j)^2+(d_k-d^*_j)^2\big],
\end{split}
\end{equation}
where $\mathcal{A}_j$ is the pool remaining after earlier recruitments;
recruited bars are excluded from the diagonal term and back-propagate
through the same circumradius machinery.
This deliberately departs from a metric-faithful $W_2$;
\S\ref{sec:results} probes its known pathology (location-blindness)
directly, and the C6 ablation shows the term is load-bearing in training.
% src: methods/topological_loss.py plan_topo_loss (greedy recruit loop)
% Round-4: greedy-vs-optimal justification, grounded in the refresh logs
% (torus profile: exactly one unreached target at every refresh, all
% seeds — results.tex C6 ¶ carries the same numbers).
Greedy rather than optimal assignment is deliberate: in every logged
regime the missing-target set is a single bar (torus: one unreached
target at every refresh, every seed; \S\ref{sec:results}), where the two
coincide; when they diverge, decreasing-persistence order spends the
pool on the most significant features first.

\subsection{Calibration instead of curriculum}
$\lambda(t)$ ramps linearly from 20\% to 50\% of training, with peak set once
by a gradient-norm ratio at the first active step:
$\lambda_{\mathrm{peak}}=\rho\,\lVert\nabla_V L_{\mathrm{photo}}\rVert /
\lVert\nabla_V L_{\mathrm{topo}}\rVert$. The single knob $\rho$ is
dimensionless ($\rho{=}0.1$ throughout); the tail error is flat across
$\rho\in\{0.03,0.1,0.3\}$, a constant-$\lambda$ ablation matches the
ramped version, and alternative ramp windows land within seed noise
(suppl.\ Table~S8) --- calibration, not scheduling, makes the knob safe.
% src: PHASE3_PLAN.md Appendix C (rho sweep) + Appendix D (C3 vs C1);
% ramp_1040/quicklook.json + ramp_3060/quicklook.json (2026-07-09)
% — window numbers now printed in suppl. Table S8 (round-4 page budget)

\subsection{Conditions}
\label{sec:conditions}
C0: baseline (photometric only). C1: $+L_{\mathrm{topo}}$. C2: $+$ a
short-range repulsion on the \emph{same} frozen samples, calibrated with the
\emph{same} $\rho$ --- the norm-matched control, the RH2 test (C2g: a
gentler radius). C3: C1 without the ramp.
(C4, a curvature-weighted variant, is deferred to follow-up work; the
numbering gap keeps condition tags comparable across studies.)
C5: C1 $+$ the best prior of our allocation study (the spread topological
field B4; suppl.\ \S A) --- the RH3 test.
C6: C1 with the recruitment term removed
(pure matching $+$ diagonal, all else identical) --- does recruitment matter
in-loop, or only for cold-start repair?
% src: PHASE3_PLAN.md Appendix B (cleanliness, revised criterion; the
% one-loss-line hook / flag-off / rho=0-divergence note trimmed 2026-07-16)
