\section{Related work}
\label{sec:related}

% Reuses only citations verified for paper 1 (refs.bib copied verbatim).
% TODO(human): prose voice pass; consider adding 2-3 newer differentiable-PH
% works after verifying their entries (red TODO below).
% 3DV pass 2026-07-13: appendix refs -> supplementary; tightened wording.
% The allocation-summary ¶ and the 3-novelty ¶ are advisor round-2 items —
% substance kept intact.
% ADVISOR REQUEST 2026-07-13 ("related works in 10 years, supporting or
% differing on DiffSoup-style reconstruction"): ¶1 rewritten as a
% four-position map of gradient-based reconstruction 2018-2026; all 16 new
% entries web-verified same day (provenance comments in refs.bib).
% Positions (i)-(iii) = the "differing" families; (iv) = the supporting
% lineage this paper builds on.

\paragraph{A decade of gradient-based reconstruction: four positions on
connectivity.}
OpenDR~\citep{loper2014opendr}, neural mesh
rendering~\citep{kato2018neural}, SoftRas~\citep{liu2019softras},
DIB-R~\citep{chen2019dibr} and nvdiffrast~\citep{laine2020nvdiffrast}
established gradient flow from pixels to geometry; reconstruction work
since splits on what that geometry should be.
(i)~\emph{Implicit fields} --- learned occupancy and signed
distance~\citep{mescheder2019occupancy,park2019deepsdf}, volume-rendered
from NeRF~\citep{mildenhall2020nerf} through NeuS~\citep{wang2021neus} to
Neuralangelo~\citep{li2023neuralangelo} --- are watertight and
topologically consistent \emph{by construction} (the strongest
counter-position to explicit soups), at field-evaluation and
extraction cost.
(ii)~\emph{Fixed-connectivity templates} deformed per
scene~\citep{wang2018pixel2mesh,hanocka2020point2mesh} keep manifoldness
but freeze topology at the template's genus.
(iii)~\emph{Differentiable iso-surfacing} re-extracts the mesh from a
volumetric scaffold at every
step~\citep{shen2021dmtet,munkberg2022nvdiffrec,shen2023flexicubes}
(Shape-as-Points bridges point soups to watertight
implicits~\citep{peng2021sap}) --- topological freedom at grid cost.
% Round-5 (advisor Related-Works doc): the 2025--26 return-to-triangles
% beat made explicit inside position (iv), with the pipeline-compatibility
% rationale; MeshSplatting verified (CVPR 2026 oral) and added as the
% connected-side counterpoint sentence below — NOT inside the soup list,
% which would misclassify it.
(iv)~\emph{Connectivity-free primitive soups} --- point
splatting~\citep{yifan2019dss}, patch atlases~\citep{groueix2018atlasnet},
Gaussians~\citep{kerbl2023gaussians} and their surface-accurate
variants~\citep{huang20242dgs,guedon2024sugar}, and a 2025--26 return to
native triangles (splatted~\citep{held2025trianglesplatting},
soft-connected~\citep{burgdorfer2025radiant}, rasterized by
DiffSoup~\citep{tojo2026diffsoup}) --- are the fastest and the easiest
to budget, but even their surface-minded members regularize
\emph{geometry} (alignment, flatness, connectivity forces), never a
measured topology.
MeshSplatting, conversely, differentiably renders
\emph{already-connected} opaque meshes~\citep{held2026meshsplatting}
--- connectivity as \emph{input}; ours stays a free soup with measured
topology.
(i) and (iii) solve topology representationally; (iv) had not addressed
it. Cut by \emph{optimization philosophy} rather than representation,
the decade holds \emph{geometry-oriented} objectives (every method
above), \emph{sampling-oriented} steering of where a fixed objective
spends primitives (below), and an empty third family ---
\emph{topology-oriented} objectives that measure the property and
differentiate through it (suppl.\ Table~S6). Prior work occupies every
pairwise intersection of the section's axes --- explicit triangles,
differentiable optimization, topology in the objective; this paper
takes the three-way one, at the soup's budget and speed.
We build on DiffSoup, whose in-loop
prune/respawn resampler is the very mechanism our allocation study
(suppl.\ \S A) steered; here the resampler stays untouched and only the
objective changes. Vertex-domain preconditioning~\citep{nicolet2021largesteps} is
complementary: it changes \emph{how} vertices move, we change \emph{why}.

\paragraph{The allocation channel: why priors alone fail.}
% Round-2 self-containment: the three findings in the MAIN text (full
% detail + tables stay in suppl. §A). src: suppl. §A sources.
% Round-4: compact main-text table added (the review's top priority —
% "RH1 cannot rest off-page"); rows condensed from suppl. Table S1.
Three findings of our allocation-channel study set this paper's problem.
(1)~Geometric metrics are topology-blind: on bisection-matched probes with
Chamfer equal \emph{by construction} ($0.916$/$0.628$/$0.977\%$ per
pair), the discriminating-dimension bottleneck separates correct from
broken topology by $35$--$40\times$ --- thin handle ${\approx}0$ vs
$.0302$, bridged merge $.0014$ vs $.0574$, punctured void $.0040$ vs
$.1385$ --- and Hausdorff$_{95}$ even \emph{prefers} the wrong candidate
in two of three cases (suppl.\ Table~S1).
% src: figures/summary.md (Phase-1 result table) = suppl. Table S1
% verbatim. Round-4 note: the key numbers sit HERE in main prose; the
% front-page matrix (Fig. 1) carries the visual form of the same point,
% so the interim compact table was folded back into this sentence.
(2)~For voids, a spread topological prior
cuts the bottleneck $33$--$53\%$ below baseline --- but a width-matched
\emph{random} control recovers most of that gain, leaving a small,
shape-dependent residual: the prior's value is carried largely by its
width. (3)~No prior shape repairs loops: the best torus arm is the
\emph{non-topological} wide control ($.0267$ vs the spread prior's
$.0316$), and the concentrated field manufactures phantom handles
($4.4$ significant loops vs the true $2$). The
composition condition C5 reuses that study's strongest field (B4;
suppl.\ \S A).

\paragraph{Persistence in learning objectives.}
Persistent homology~\citep{edelsbrunner2002persistence} with its stability
guarantees~\citep{cohensteiner2007stability} has been used as a trainable
signal via differentiable sorting of filtration
values~\citep{gabrielsson2020topologylayer}, for topology-preserving image
segmentation~\citep{hu2019topopreserving,clough2022topoloss}, point-cloud
and shape optimization and 3D volume
prediction~\citep{bruel2020toporecon,poulenard2018shapematching,waibel2022topological},
and diagram-metric optimization with convergence
analysis~\citep{carriere2021optimizing}; barcode-valued maps have since been
given a general differential framework~\citep{leygonie2022framework}, and the
per-step cost of persistence-guided descent attacked
directly~\citep{nigmetov2024bigsteps} --- yet never as the objective of
an \emph{explicit triangle-based} reconstruction.
% Round-5 (advisor Related-Works doc): the requested bridge — PH broadly
% applied, the explicit-triangle objective slot open — adapted, not
% pasted; merged into the survey sentence for the page budget.
Nearest our setting, concurrent work brings persistence to
reconstruction at other, complementary entry points: an image-space barcode term for
Gaussian splatting~\citep{shen2025topologygs}, homology-guided camera
placement~\citep{gao2026homology} or a genus-matched
template~\citep{gao2025genus} in mesh inverse rendering, and a
connected-component constraint on implicits fitted to point
clouds~\citep{jignasu2024stitch}.
% Round-4: the suppl. §D.7 three-criteria partition promoted here — it is
% the per-method infeasibility argument the venue expects in place of a
% direct benchmark (inputs/outputs/cost regimes do not align).
Partition our setting's three requirements --- (i)~a topological quantity
of the evolving \emph{surface} measured differentiably in the objective,
(ii)~image-based training through a renderer, (iii)~a fixed primitive
budget --- and each neighbour satisfies at most one (suppl.\ \S D.7): the
image-space barcode term never measures the surface (i); the point-cloud
pipeline forgoes renderer and budget (ii, iii); the template line
\emph{assumes} the genus we measure (i). A head-to-head benchmark is
therefore ill-posed (inputs, outputs, and cost regimes differ); our
evaluation instead pits the loss against norm-matched controls through
its own channel (\S\ref{sec:setup}).
% Round-5 (advisor Related-Works doc): nearest connectivity-side work
% named and contrasted — soft connectivity forces are the heuristic the
% objective replaces (burgdorfer2025radiant: round-3-verified,
% arXiv:2505.23642). Kept OUTSIDE the three-requirement partition above,
% which scopes over the persistence neighbours only.
The nearest soup-side neighbour regularizes coherence with \emph{soft
connectivity forces}~\citep{burgdorfer2025radiant} --- local continuity
pressure, no measured topology; we replace that heuristic with the
measurement. Our backward pass is in the persistence family
--- unlike work that back-propagates through filtration-value sorting or
barcode coordinates in general position, we re-express each paired
simplex's birth/death as a closed-form circumradius --- with three
pipeline-specific ingredients.
(i)~\emph{Alpha-complex pairing over $\alpha{\times}$area surface samples}
of a live triangle soup --- not image grids or raw clouds --- couples
the diagram to the geometry actually being trained.
(ii)~A \emph{Gabriel-exactness gate}: the circumradius re-expression
equals the filtration value only for Gabriel simplices, so we verify it
per pairing (relative $10^{-6}$) --- certifying the frozen gradient
rather than assuming it.
(iii)~\emph{Recruitment for missing features}: optimal partial matching
has zero gradient exactly when a target feature is entirely absent --- the
worst failure mode --- so we trade metric fidelity for a restored descent
direction (\S\ref{sec:method}), quantifying both its pathology and its
necessity (\S\ref{sec:results}).
% Leygonie/Nigmetov entries web-verified 2026-07-09 (Springer/arXiv); see
% refs.bib provenance comments.

\paragraph{Topology-aware geometry processing.}
Topology-aware reconstruction and repair predate differentiable
pipelines~\citep{sharf2007topologyaware,ju2004repair,attene2010repair}, and
budgeted remeshing has a long
lineage~\citep{hoppe1996progressive,garland1997qem,alliez2003anisotropic,botsch2004remeshing,dunyach2013adaptive};
our contribution is orthogonal --- the budget mechanics stay DiffSoup's,
only the training signal becomes topology-aware.
% Round-5 (advisor Related-Works doc): the requested three-axis gap
% recut lives in ¶1's closing sentence (merged with the philosophy cut
% to pay the page budget); ¶4 stays the classic-processing coda.
