\section{Results}
\label{sec:results}

% Every number: % src: topo3/report/results.json (regenerate via
% experiments/topo_loss_report.py) unless marked otherwise.
% 3DV pass 2026-07-13: tab:main/tab:gen/fig:series -> full-width floats;
% fig:tails + fig:gen + the stage-3a gate (prose + fig:toys) moved to
% supplementary §C (Figs. S2-S4) with captions verbatim; pot ¶ merged into
% the observations ¶ (they repeated each other); prose + captions
% tightened. Every number and % src kept.

A gate first: the loss alone (no photometric term, Adam on points)
repairs all four controlled probes --- a punctured void closes
($8.8\times$), a spurious handle dies at the diagonal, mis-spaced
components separate ($1229\times$), a bridged merge with \emph{no}
significant live H0 bar (Lemma~\ref{lem:zerograd}'s case) is severed
($450\times$, $\beta_0\,1{\to}2$); the predicted location-blindness did
not materialize (suppl.\ \S C, Fig.~S5). Raw-cloud sanity toys;
multipliers not comparable to training.
% src: figures/phase3_toy/t{1..4}.json (PHASE3_PLAN.md Appendix A)

\subsection{The C-matrix}
\label{sec:cmatrix}
Table~\ref{tab:main} reports tail bottleneck (mean$\pm$sd over seeds) in each
shape's discriminating dimension; Chamfer in the text where it matters;
the full training trajectories for the two decisive shapes are in
suppl.\ Fig.~S6 (line = seed mean, band = seed min--max).
% Round-4: fig:series moved to suppl. §C (Fig. S6) to fund the teaser +
% qualitative-matrix page budget; caption carried verbatim.

\begin{table*}[t]
\centering\small
\caption{Tail bottleneck to target (mean$\pm$sd; lower is better);
$\times$ = vs C0; verdict: C1 $<$ C0 and $<$ every control at Chamfer
parity. Seeds: 5 (sphere, torus, double torus --- saturated
identically), 3 (cube, two spheres). C6 = no recruitment.}
\label{tab:main}
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lllllll}
\toprule
shape (dim) & C0 & C1 loss & C2 control & C3 no-ramp & C5 loss+prior & C6 no-recruit \\
\midrule
% src: topo3/report/results.json (C6 column added round 4: sphere
% .016564+-.000498, torus .041147+-.003824)
sphere (H2)      & .0623$\pm$.0063 & \textbf{.0156$\pm$.0013} (4.0$\times$) & .1372 (worse) & .0151$\pm$.0004 & \textbf{.0082$\pm$.0007} (7.6$\times$) & .0166$\pm$.0005 \\
cube (H2)        & .0582$\pm$.0046 & \textbf{.0074$\pm$.0011} (7.9$\times$) & .1346 ($\beta_2\,1{\to}0$) & --- & --- & --- \\
torus (H1)       & .0424$\pm$.0031 & \textbf{.0180$\pm$.0016} (2.3$\times$) & .0457 ($\beta_1\,2{\to}1$) & .0155$\pm$.0015 & \textbf{.0133$\pm$.0004} (3.2$\times$) & .0411$\pm$.0038 \\
two spheres (H0)& .0065$\pm$.0008 & .0007$\pm$.0002 (9.5$\times$)          & .0008 (ties C1) & --- & --- & --- \\
double torus (H1)& .0264$\pm$.0000 & .0264$\pm$.0000 (saturated)           & .0264$\pm$.0000 & --- & --- & --- \\
\bottomrule
\end{tabular}
\end{table*}

\paragraph{Voids (H2): topology-specific, large.} Sphere $4.0\times$
($16.2\sigma$ Welch vs C0, five seeds), cube $7.9\times$ (three seeds:
ranges disjoint, C1 $.0064$--$.0086$ vs C0 $.0551$--$.0636$), both at
equal-or-better Chamfer (sphere $0.46$ vs $0.50$; cube $0.92$ vs $1.05$).
% src: topo3/report/results.json cube tails C1 .006370/.007109/.008617,
% C0 .055129/.056018/.063560 (18.5-sigma print retired at n=3, round 4)
The
norm-matched control is \emph{destructive}, not merely weaker: it pins
the void's death at its repulsion equilibrium ($2.2\times$ worse,
seed-identical) and erases the cube's void outright; no tested strength
reproduces any gain --- C1 $<$ C0 $<$ C2g $<$ C2
(\S\ref{sec:limitations}; spread: suppl.\ Fig.~S3; mechanism:
\S\ref{sec:discussion}).
% src: topo3/report/results.json (sphere: C2g 0.0972+-0.0169; the
% number itself is printed in the Limitations "Control strength" item)

\paragraph{Loops (H1): the channel's distinguishing result.} No prior
shape won this class (suppl.\ \S A). The loss cuts torus
H1 error $2.3\times$ with \emph{zero phantom handles in all five seeds}
(count trajectories flat at the true $\beta_1{=}2$,
suppl.\ Fig.~S4), while the control collapses one loop by step
${\sim}800$ and never recovers it. Metric pressure
on birth/death values suits 1-D features in a way budget allocation does
not (mechanism: \S\ref{sec:discussion}).

\paragraph{Components (H0): not topological --- third channel, same answer.}
C1 reduces the (already small) baseline error $9.5\times$, but the
generic control matches it (.0008 vs .0007); the increment appears only
in Chamfer ($0.54$ vs $1.94$). Three channels agree: $\beta_0$ is
carried by shell count and separation.

\paragraph{Composition (C5).} Loss $+$ spread prior is best everywhere
tested --- sphere $.0082$ ($7.6\times$ vs C0, ${\approx}2\times$ vs C1 alone,
${\approx}4.4\times$ below the prior alone
% src: h2_unified/report/results.json (B4 sphere .0357; .0357/.0082 = 4.37)
), torus $.0133$ --- with the best sphere Chamfer (torus $.502$,
trailing only C3's $.495$). Allocation and value-tuning are
complementary, not substitutes.
% src: topo3/report/results.json (chamfer_mean per arm)

\paragraph{Stretch case (genus 2): saturated, honestly.} Every double-torus
arm returns exactly $.0264$: the two tube loops are unexpressable at
feasible budgets --- the \emph{representation's} ceiling, not the
channel's (suppl.\ \S B; floors: \S\ref{sec:discussion}).
% src: topo3/report/results.json; detail moved to suppl. §B (2026-07-16)

\paragraph{Ablations.} The $\rho$-response is flat over a $10\times$
range (suppl.\ Table~S8); C3 (no curriculum)
$\approx$ C1 on both decisive shapes \emph{and} off the analytic family
(bob: C3 $.0171{\pm}.0026$ vs C1 $.0208{\pm}.0008$).
% src: PHASE3_PLAN.md Appendix C (rho mini-sweep, sphere seed 0, tail =
% last 3 dumps — values now printed in suppl. Table S8);
% topo3/report/results.json (bob C3, 3 seeds, 2026-07-09)
Overhead, honestly: the loss adds a \emph{fixed}
${\approx}46$\,s per 2{,}500-step run (refresh $174$\,ms at $M{=}2048$
every 10 steps) --- more than the $33$--$41$\,s photometric baseline on
these deliberately tiny scenes, but the cost scales $O(M\log M)$ per
refresh, not with scene or image complexity, so its share of a run whose
photometric cost is $T$ is $46/(T{+}46)$ --- an arithmetic projection, ${\approx}7\%$ of a ten-minute run ---
independent of image count and resolution; no parameters are added.
% src: PHASE3_PLAN.md Appendices B-D; wall-clock from dump mtimes
% (sphere C0 33s / C1 79s; torus C0 41s / C1 88s, 5 seeds each)

\paragraph{Recruitment is load-bearing for loops (C6).} Removing the
recruitment term (pure matching $+$ diagonal) collapses the torus win
entirely: $.0411\pm.0038$, statistically baseline ($0.6\sigma$ from C0,
$12.4\sigma$ worse than C1) at baseline Chamfer, though both loops
still \emph{exist} ($\#$sig H1 $= 2$ every seed; suppl.\ Fig.~S6). The
logs locate the mechanism: one torus loop is significant at every
refresh, the other chronically \emph{below} threshold (all five seeds,
all 200 refreshes) --- recruitment is the second loop's sole gradient
path; on this class it \emph{is} the mechanism, not a cold-start
device. On the sphere it never fires (0 of 200, all seeds): C6 is
loss-identical to C1, and the $1.5\sigma$ tail gap ($.0166$ vs $.0156$)
measures pure run-to-run noise (\S\ref{sec:limitations}).
% src: topo3/report/results.json (C6 arms) + topo3/_train/*_C6_*/
% topo_loss_log.json ("refreshes": per-refresh matched/target_unreached)

\paragraph{Sensor-noise stress (C7/C7h).} With the plan's cloud
perturbed at every refresh ($\sigma = 0.5\%$/$1\%$ of the diagonal
${\approx}0.5$/$1.0$ spacings, \S\ref{sec:setup}), the loss degrades
gradually and never harms: torus $.0210{\pm}.0018$ / $.0259{\pm}.0022$
($2.0\times$/$1.6\times$ below baseline, vs C1's $2.3\times$), sphere
$.0235{\pm}.0009$ / $.0328{\pm}.0013$ ($2.7\times$/$1.9\times$, vs
$4.0\times$), at equal-or-better Chamfer, correct counts in every run
(suppl.\ Fig.~S6), zero Gabriel failures in all $2{,}400$ noisy
refreshes (internals: suppl.\ \S B).
% src: topo3/report/results.json (C7/C7h arms) + topo3/_train/*_C7*/
% topo_loss_log.json ("refreshes": n_gabriel_fail / recruit /
% target_unreached per refresh; lam_peak per run)

\subsection{Generality: external genus-known meshes}
\label{sec:generality}
Everything above is on one analytic family, so we repeat the verdict
protocol --- C0/C1/C2, three seeds, class budgets, $\rho$ and ramp
unchanged, density fixed by the floor rule --- on the eight external
genus-known meshes of \S\ref{sec:setup}'s two groups, with no
per-shape tuning.
% src: topo3/report/results.json (all eight aggregates + verdicts);
% tom-yum pot = internal shape key kinkin (quicklook cross-checked);
% group wave 2026-07-19: GROUP_WAVE_PLAN.md + group_wave_stats.json
Seven of eight \textbf{pass} the verdict rule at equal-or-\emph{better}
Chamfer (Table~\ref{tab:gen}; original-wave trajectories in suppl.\
Fig.~S2); the eighth, rocker-arm, is a \emph{no-headroom null}, not a
counterexample (obs.~5). The control replicates its failure modes: all
five voids erased ($\beta_2\,1{\to}0$, every seed; the pot at
$1.36\times$ worse Chamfer), a bob loop collapsed ($\beta_1\,2{\to}1$),
eight's missing pair left missing.

\begin{table}[t]
\centering
% Round-4 statistics fix: the n=3 Welch sigmas (6.1/26.1/35.7/80.1) are
% retired — near-zero sd at three seeds makes sigma meaningless; verdicts
% rest on disjoint per-seed ranges (suppl. Table S7). Round-4 fit pass:
% single-column float; the all-PASS verdict column folded into the
% caption.
% Round-5 group wave (2026-07-19): grouped rows + the advisor's group
% means; pre-registered statistic = mean +- sd over member shapes of
% UNROUNDED per-shape reductions (GROUP_WAVE_PLAN.md;
% scripts/group_wave_stats.py -> output/group_wave_stats.json).
\caption{Generality wave, two observable-class groups (mean$\pm$sd, 3
seeds; $\times$ = vs C0; \emph{group mean} $\pm$ sd over member shapes,
pre-registered). Seven of eight \textbf{pass} Table~\ref{tab:main}'s
verdict rule; rocker-arm is a no-headroom null (counts intact in every
arm). *seed-exact unmatched-target pins (\S\ref{sec:discussion}).
Per-seed values: Table~S7; provenance, certificates, staircases:
suppl.\ \S G.}
\label{tab:gen}
\scriptsize
\setlength{\tabcolsep}{1.5pt}
\resizebox{\linewidth}{!}{%
\begin{tabular}{llll}
\toprule
shape & C0 & C1 loss & C2 control \\
\midrule
% src: topo3/report/results.json tails/nsig_final; pot row cross-checked
% against topo3/quicklook.json (kinkin, 2026-07-10)
bob          & .0426$\pm$.0012 & \textbf{.0208$\pm$.0008} (2.1$\times$) & .0437 ($\beta_1\,2{\to}1$) \\
rocker-arm   & .0083$\pm$.0004 & .0087$\pm$.0015 (0.96$\times$) & .0113 (counts intact) \\
eight (genus 2) & .0249$\pm$.0000* & \textbf{.0162$\pm$.0017} (1.5$\times$) & .0249 (2 of 4 loops) \\
% src: group_wave_stats.json loop group mean_reduction 1.5153 sd 0.5480
\emph{loop group mean (H1, 3)} & & \textbf{1.52$\times$}$\,\pm\,$0.55 & \\
\midrule
spot         & .0521$\pm$.0081 & \textbf{.0237$\pm$.0000} (2.2$\times$) & .0652 ($\beta_2\,1{\to}0$) \\
fandisk      & .0538$\pm$.0022 & \textbf{.0052$\pm$.0008} (10.4$\times$) & .0571 ($\beta_2\,1{\to}0$) \\
tom-yum pot  & .0221$\pm$.0003 & \textbf{.0057$\pm$.0002} (3.9$\times$) & .0223 ($\beta_2\,1{\to}0$) \\
armadillo    & .0511$\pm$.0000* & \textbf{.0124$\pm$.0001} (4.1$\times$) & .0511 ($\beta_2\,1{\to}0$) \\
horse        & .0408$\pm$.0000* & \textbf{.0109$\pm$.0022} (3.8$\times$) & .0408 ($\beta_2\,1{\to}0$) \\
% src: group_wave_stats.json void group mean_reduction 4.8677 sd 3.1741
\emph{void group mean (H2, 5)} & & \textbf{4.87$\times$}$\,\pm\,$3.17 & \\
\bottomrule
\end{tabular}}
\end{table}

% Round-4: fig:tomyum moved to suppl. §C (Fig. S7) — the pot's showcase
% render returns to main inside the round-4 qualitative matrix.
Observations.
(1)~\emph{The H1 claim survives the family change}: bob halves loop
error, zero phantom handles every seed (suppl.\ Fig.~S8), replicating
torus.
(2)~\emph{The genus-2 ceiling was representational, not topological}:
on the fat CGAL eight, baseline and control read only two of the four
certified loops (value pinned at the unmatched bound, seed-exact); the
loss \emph{restores the missing pair} ($\#$sig $4/3/4$) and cuts value
error $1.5\times$ at $0.89\times$ Chamfer --- the double torus's
saturation (\S\ref{sec:cmatrix}) was the representation's ceiling, not
genus-2's.
% src: results.json eight tails/nsig (C0 [2,2,2], C1 [4,3,4], C2
% [2,2,2]); pin: eight_src_cert.json staircase top_H1@20000 .0498/2;
% chamfer 0.7212/0.8062 = 0.894
(3)~\emph{Sharp-featured CAD benefits most}: fandisk's $10.4\times$
exceeds every analytic-family effect.
(4)~\emph{Baselines can pin like controls}: armadillo and horse read
the correct void count but pin its value at the same bound
($.0511$/$.0408$, seed-exact); the loss unpins both
($4.1\times$/$3.8\times$) at better Chamfer.
% src: results.json armadillo/horse tails; cert staircases top_H2@20000
% .1022/2, .0817/2; chamfer 0.7442/0.8469, 0.7944/0.8812
(5)~\emph{No headroom, no effect --- and no harm}: rocker-arm's one
measurable loop is already baseline-correct ($.0083$ vs the
$.04$--$.05$ band): \S\ref{sec:setup}'s imperfection criterion is unmet
and the loss stays inert ($0.96\times$, parity $1.00$, counts intact
everywhere) --- the method's honest boundary.
% src: results.json rockerarm (C0 .008298+-.000435, C1 .008666+-.001533,
% chamfer parity 1.0029; nsig [1,1,1] all arms)
(6)~\emph{Count-level limits persist on sub-budget organic detail}:
spot's \emph{baseline} makes a phantom void in two of three seeds
($\#$sig $=2,1,2$); the loss clears two, retains one ($1,2,1$) ---
value repair reliable, count repair not guaranteed; its C1 tail pins at
$.0237$ (sd $0$), a milder floor of the double-torus kind.
% src: topo3/report/results.json (nsig_final; tails)
\emph{And the pipeline reaches raw artist content}: the tom-yum pot
(Fig.~\ref{fig:matrix}; suppl.\ Fig.~S7), a raw Blender export
certified rather than constructed and bundled at $M{=}8192$ by the same
pre-registered rule (\S\ref{sec:setup}), cuts the void's value error
$3.9\times$ at Chamfer parity ($0.99\times$), the study's tightest seed
spread (C1 $.0056$--$.0059$ vs C0 $.0218$--$.0223$); the control
destroys the void ($\beta_2\,1{\to}0$, all seeds), pinned at the
unmatched bound ($.0223$) --- count, not value, is the verdict. Two caveats transfer: its baseline already reads the
correct count (a value-accuracy win, like bob), and ground truth is
certified, not constructed.
% src: topo3/quicklook.json (kinkin, cross-checked vs results.json);
% pin bound + staircase: PHASE3_STATUS.md Post-3e
Group statistic (pre-registered): loops $1.52\times\pm0.55$ (2/3 pass),
voids $4.87\times\pm3.17$ (5/5); the passing wave spans
$1.5$--$10.4\times$ (suppl.\ \S G).
% src: output/group_wave_stats.json (per-group mean/sd/min/max;
% external span 0.96-10.38 including the null)
