\section{Experimental setup}
\label{sec:setup}

% Budgets/headroom carried over from the allocation study (suppl. §A).
% TODO(human): voice pass. (The old "venue-dependent compression" item was
% done in the 3DV pass 2026-07-13 — only the voice read remains.)
% 3DV pass 2026-07-13: appendix ref -> supplementary; ingest prose
% tightened. Every number kept verbatim with its % src.

\paragraph{Scenes and budgets.} Synthetic COLMAP scenes with analytically
known topology, rendered from 24--72 Fibonacci-sphere viewpoints at
200--256\,px (every-8th test holdout $\Rightarrow$ 21--63 training
views): sphere and cube
(void class H2, $\beta{=}(1,0,1)$, budget $N{=}1200$), torus (loop class H1,
$(1,2,1)$, $N{=}700$), two spheres (component class H0, $(2,0,2)$, $N{=}700$),
and genus-2 double torus as a stretch case ($(1,4,1)$, $N{=}2000$). Budgets
are the allocation study's headroom points (suppl.\ \S A) --- tight
enough that baseline topology is imperfect.
% src: PHASE2B_CROSSOVER.md (headroom pilot); PHASE3_PLAN.md §6
% Round-5 group wave (2026-07-19): externals expanded 4 -> 8 in the two
% observable-class groups the advisor asked for; selection/protocol
% pre-registered (GROUP_WAVE_PLAN.md; suppl. §G has provenance,
% certificates, staircases, and the honest exclusions).
Eight external, genus-known meshes probe generality beyond the analytic
family, organized as the study's two observable classes
(\S\ref{sec:generality}): a \emph{loop group} --- \emph{bob} (genus 1),
\emph{rocker-arm} (1), the CGAL \emph{eight} (2) --- and a \emph{void
group} --- \emph{spot}, \emph{fandisk}, \emph{armadillo}, \emph{horse}
(all genus 0), the \emph{tom-yum pot} (3, below); sources, licenses and
certificates: suppl.\ \S G. Seven pass the mesh certificate below
as-is; five further
candidates failed it (open boundaries, non-manifold junctions) and were
excluded, never repaired. Membership, observable class and density were fixed by the
pre-registered staircase rule \emph{before} any training (rocker-arm:
torus precedent, H1-restricted at $M{=}2048$; eight: the pot's, full
$(1,4,1)$ bundle at $M{=}8192$). Scenes: the same pipeline (48 views,
200\,px), unchanged class budgets (trio pilot: tails $.043$--$.054$,
in the analytic band).
% src: output/synth/{spot,bob,fandisk} scene builds; pilot tails = the
% seed-0 entries of the topo3/report/results.json C0 aggregates
% (.0429/.0429/.0540 — the pilot's 2026-07-09 quicklook.json was later
% overwritten by the tomyum/kinkin quicklook, so cite results.json);
% bundles topo3/fields/{spot,bob,fandisk}_diag.npz
The fourth, the \emph{tom-yum pot} (Fig.~\ref{fig:matrix}, bottom row;
aluminium showcase: suppl.\ Fig.~S7), probes the step
every real deployment needs: a raw artist-authored mesh
rather than a clean benchmark. As exported from Blender it is textbook
triangle soup (9 overlapping open shells, 232 non-manifold junction
edges) with no trustworthy ground-truth topology --- the same defect
that excluded ShapeNet. An ingest-and-certify pass (exact weld,
offset-solidification into a thin closed shell, then \emph{certification
by measurement}: exact simplicial homology cross-checked against
per-component Euler characteristics, both independent of the loss's
sampled alpha complex; watertightness; bit-identical rebuilds) yields
one body of genus 3, $\beta{=}(1,6,1)$ exact,
% Round-5 (revision list, minor "circular certification" item): the
% independence of the certificate computations made explicit.
% src: scripts/make_kinkin_asset.py steps 5b (euler_number -> genus,
% chi cross-check) + 5c (gudhi exact mesh homology, asserted equal);
% the alpha-complex staircase is the separate step-7 preflight.
rendered and budgeted as a void-class shape ($N{=}1200$). This ingest
pipeline is a test fixture here, not a claimed contribution. Where the
other externals arrive clean enough to certify as-is, the pot's
topology exists only \emph{after} ingest --- a controlled real-world
proxy, not a blind test: what real scans will need (mechanics and
challenge inventory: suppl.\ \S B, Table~S4).
% Round-5 fixes here: (a) REPAIRED a comment-swallowed sentence-start
% ("rendered and budgeted…" had been appended to the src-comment line
% by the earlier cert edit — the fragment was missing from the PDF,
% incl. commit 3e829aa); (b) the construction-vs-certificate contrast
% went stale once the whole wave is certificate-verified (suppl. §G).
% Round-5 (revision list, Critical row 1): the advisor's "controlled
% real-world proxy" label adopted verbatim here; the abstract's scope
% caveat and §7's "Synthetic, single-machine" item already carry the
% same honesty.
% Round-4 trim: raw-soup counts (6,318 verts, quad/12-gon mix, 50
% boundary edges, 804 dropped shells, distance-field/marching-cubes
% detail, 190,990 certified verts) now live ONLY in suppl. Table S4 —
% numbers unchanged there, % srcs intact.
% naming: rendered as tom-yum pot; the asset, tags and files are named
% kinkin internally (do not confuse with the constructed CSG pot, which is
% NOT in this paper).
% src: output/synth/_meshes/kinkin_src_cert.json (components/genus/betti/
% verts + raw-soup counts) + scripts/make_kinkin_asset.py (ingest+certify,
% bit-identical rebuild check); scene: output/synth/kinkin
The pot also exercises the floor rule of \S\ref{sec:discussion}
prospectively: every H1 loop of the certified shell sits below the
$6\,r_{\mathrm{med}}$ floor at every working $M$, while the narrow flue
mouth caps at small $\alpha$ --- an H2 void on the pot axis, first
clearing the floor at $M{=}8192$ (margin $1.20$--$1.23\times$, five
sampling seeds). The pre-registered rule builds the bundle there (a
floor-rule density deviation later repeated by eight; the loss samples
at the same density): exactly one significant bar, stable to
re-sampling (${\sim}10^{-5}$).
% Round-5 accuracy fix: "the study's only protocol deviation" became
% false when eight also bundled at 8192 (suppl. §G) — rephrased.
% src: PHASE3_STATUS.md Post-3e (make_kinkin_asset.py preflight staircase:
% (1,0,0)@2048/4096 -> (1,0,1)@8192/20000; margins recomputed live by
% scripts/make_kinkin_brief_docx.py); floor rule: experiments/density_bound.py

\paragraph{Protocol.}
% Round-4: implementation provenance made explicit (review item 148-154).
All conditions share one implementation: C0 is the unmodified DiffSoup
trainer; the loss enters as one additive term, and disabled it executes
none of the added code (no renderer or resampler edits).
% src: PHASE3_STATUS.md "What was built" (ONE loss line after reg_normal;
% flag-off executes zero Phase-3 code; no renderer/core edits)
2{,}500 steps; resampling every 100 steps (untouched);
$\rho{=}0.1$, ramp $0.2{:}0.5$, refresh $K{=}10$, $M{=}2048$; decisive shapes
(sphere, torus) and the double torus run 5 seeds (the former with the
full condition set incl.\ C6), replication shapes (cube, two spheres)
3 seeds (C0/C1/C2).
% Round-4: double torus raised 2 -> 5 seeds (2026-07-17, same
% pre-registered flags, gudhi 3.11 verified equivalent); every arm
% still returns exactly .026372 in every seed — the saturation claim
% now carries five seeds. src: topo3/report/results.json A sensor-noise stress test
(C7/C7h $=$ C1 $+$ noise; decisive shapes, 3 seeds) perturbs the cloud
the loss \emph{plan} sees with i.i.d.\ Gaussian noise at every refresh:
the plan decides on the noisy observation, the pulls apply to the true
live positions.
% src: methods/topological_loss.py TopoLossState(noise=...); r_med at 2048:
% experiments/density_bound.py (torus .0104, i.e. 1% of diag ~= 1 spacing)
The torus restricts the loss to H1 (its own void bar is sub-threshold at
$M{=}2048$ --- a loss must not act on features it cannot reliably
measure); the double torus likewise targets its two measurable loops of
four. The generality wave runs C0/C1/C2, three seeds, each shape at its
staircase-fixed bundle (bob: full loops$+$void; rocker-arm: H1 only;
eight and the pot: $M{=}8192$).
% src: PHASE3_PLAN.md Appendix C (bundle preflights) + PHASE2B (2/4 loops);
% spot/bob/fandisk bundle preflights 2026-07-09 (H bars printed at build)

\paragraph{Metrics and verdict.} Primary: bottleneck to the target
diagram in the discriminating dimension, tail-averaged (last three
dumps, steps $2{,}300$--$2{,}500$), seed-averaged (20k eval samples,
target-normalized).
% src: experiments/topo_loss_report.py (--tail default 3; dumps every 100)
Secondary: significant-feature counts (phantom check) and Chamfer parity.
% Round-4 reporting policy (statistical-hygiene items): sigma only at
% n=5; ranges at n=3; descriptive at n=2.
Reporting policy: mean$\pm$sd over seeds; Welch $\sigma$ only where both
arms ran five seeds; three-seed verdicts rest on disjoint per-seed
ranges (suppl.\ Table~S7).
\emph{Verdict rule}: C1 must beat C0 \textbf{and} every norm-matched
control at Chamfer parity --- a topological win a dumb regularizer can
match is not one. \emph{Chamfer parity}, formally: the passing arm's
seed-mean Chamfer must satisfy
$\mathrm{Ch}(\mathrm{C1})\le 1.15\,\mathrm{Ch}(\mathrm{C0})$; the $15\%$
headroom absorbs seed variance and nondeterminism
(\S\ref{sec:limitations}) and never did any work --- every
passing arm's Chamfer was at or below baseline.
% src: experiments/topo_loss_report.py PARITY_MULT = 1.15; per-arm
% chamfer_mean in topo3/report/results.json (all passing arms <= C0)
% src: experiments/topo_loss_report.py verdict rule; PHASE3_PLAN.md App. B
