% ===== 3. Method (DRAFTABLE from code) ======================================
% Grounded in: methods/topo_importance.py and methods/topo_resampling.py.
% No experimental numbers in this section.
\section{Topology-aware adaptive resampling}
\label{sec:method}

The method has two offline-then-online parts: a \emph{precomputed importance
field} distilled once from the target, and a \emph{resampling policy} that
consumes the field inside the training loop. Neither touches the renderer, and
both are budget-neutral.

\subsection{The importance field}
\label{sec:method-field}
% src(design): methods/topo_importance.py (build_importance_field, _localize_features)
Let $P$ be a seeded surface sample of the target. We compute its persistence
diagram~\cite{edelsbrunner2002persistence} with a GUDHI~\cite{maria2014gudhi}
alpha complex~\cite{edelsbrunner1994alphashapes} and keep the \emph{significant}
features---intervals whose lifetime exceeds a self-calibrating multiple of the median
nearest-neighbour spacing. Each significant feature $f$ of homological dimension
$d_f$ is \emph{localized} in space using the persistence pairs together with the
alpha-complex point coordinates: an H1 loop is located by the edge/triangle that
births/fills it, an H2 void by the tetrahedron that fills it (which sits on the
bounding shell), and an H0 component by the edge that merges it. This yields, per
feature, a set of 3D kernel centers $\{c\}_f$ on the target surface.
% TODO(human): if you include a localization figure, add it here.

Each feature contributes an isotropic Gaussian splat, and the field is evaluated
on a dense surface sample $Q$ (an ``option-B surface splat'') so that importance
is tied to the manifold the soup must reproduce:
\begin{equation}
  r(q) \;=\; \sum_{f}\; w_f \sum_{c\in\{c\}_f}
            \exp\!\Big(-\tfrac{\lVert q-c\rVert^2}{2\,\sigma_f^2}\Big),
  \qquad q\in Q,
  \label{eq:splat}
\end{equation}
with per-feature weight $w_f$ equal to the (normalized) persistence and radius
$\sigma_f$ set to the feature's own death scale---its alpha circumradius
$\sqrt{\text{death}_f}$, clamped to a spacing floor and a bounding-box-fraction
ceiling. The per-dimension fields are normalized to $[0,1]$ and a query point
takes the value of its nearest sample in $Q$.

\paragraph{Concentrate$\leftrightarrow$spread scale.}
% src: methods/topo_importance.py (_localize_features: sigma=s*sigma_base, w=life/s**2)
We add a single scale $s\!\ge\!1$ that widens every kernel while preserving the
injected surface mass: $\sigma_f \to s\,\sigma_f$ and
$w_f \to w_f/s^2$. Because a 2D Gaussian's surface integral scales as
$w\sigma^2$, the total importance mass $\sum_f w_f\sigma_f^2$ is invariant in $s$;
$s$ changes only the spatial \emph{spread} of the prior, not how much of it there
is. By construction $s\!=\!1$ is the \emph{concentrated} field and is numerically
identical to the field used without the knob.
% TODO(human): the mass-invariance and s=1 identity were checked empirically
% (Sigma w sigma^2 constant across s; s=1 field byte-identical to baseline field).
% Cite that check or move it to the appendix if you want it in-paper.

\subsection{The resampling policy}
\label{sec:method-policy}
% src(design): methods/topo_resampling.py (TopoResamplePolicy: keep_map, densify_and_rebalance)
The baseline resampler periodically (i) \emph{prunes} the least screen-visible
triangles and (ii) \emph{respawns} triangles by subdividing large screen-space
edges. Our policy biases both with the field via one strength knob
$\lambda_{\text{topo}}$, and keeps the triangle count $N$ fixed:

\begin{itemize}
  \item \textbf{Prune protection (keep-map).} Faces are ranked by
  $\text{visibility} + \lambda_{\text{topo}}\,\kappa\,\mathrm{imp}$, where
  $\mathrm{imp}$ is the field value at the face centroid and $\kappa$ a
  visibility-scale factor; the same number of faces is removed as in the baseline,
  so important faces are protected without changing the budget.
  \item \textbf{Budget-neutral densification.} After the unchanged baseline split,
  we add up to $\mathrm{cap}=\lambda_{\text{topo}}\!\cdot\!c\!\cdot\!N$ extra
  subdivisions of the highest-importance visible faces, then remove exactly
  $\mathrm{cap}$ faces with the protective keep-map. Net face count is unchanged;
  the effect is a pure trade of low-importance triangles for high-importance ones.
\end{itemize}

When $\lambda_{\text{topo}}=0$ both levers are bypassed and the policy reduces
\emph{exactly} to the baseline resampler (bit-identical resampling decisions);
this is condition B0 below. The policy is field-agnostic: swapping which field it
reads (concentrated topological, spread topological, or a non-topological control)
gives the different experimental conditions of Section~\ref{sec:setup} without any
other change.
% TODO(human): confirm you want the keep-map / densify formulae at this level of
% detail (vs. an algorithm box). Symbols kappa, c are protect-scale and
% split_cap_frac in the code.
