% ===== 4. Experimental setup (DRAFTABLE) ====================================
\section{Experimental setup}
\label{sec:setup}

\paragraph{Conditions.}
All conditions are identical in seed, target scene, triangle budget $N$,
optimizer, step count, and schedule; they differ \emph{only} in the resampling
importance (Section~\ref{sec:method-policy}).
\begin{itemize}
  \item \textbf{B0} --- baseline resampler ($\lambda_{\text{topo}}=0$); the control.
  \item \textbf{B1} --- topological field applied \emph{once at initialization} only;
        tests whether init-time bias survives (it does not).
  \item \textbf{B2} --- topological field applied in-loop, \emph{concentrated}
        ($s\!=\!1$); the proposed method.
  \item \textbf{B3} --- a \emph{non-topological} importance field (random blobs at the
        default \emph{narrow} width) applied in-loop; with B2 it tests whether narrow
        topological placement beats narrow non-topological densification.
  \item \textbf{B4} --- topological field applied in-loop but \emph{spread}
        ($s\!>\!1$); identical to B2 except for the concentrate$\leftrightarrow$spread
        scale. \emph{Results in Sections~\ref{sec:results-h2} and \ref{sec:results-crossover}.}
  \item \textbf{B5} --- a \emph{non-topological} field \emph{width-matched} to B4
        (random blobs widened to B4's kernel $\sigma$, mass-preserved); with B3 and B4
        it separates topological \emph{placement} from kernel \emph{width}
        (H2 shapes; Section~\ref{sec:results-h2}).
\end{itemize}

\paragraph{Synthetic targets with known topology.}
Targets are analytic shapes rendered into multi-view scenes so the ground-truth
surface (and hence its Betti numbers) is exact. Each shape's Betti numbers are
recovered by the alpha-complex pipeline before any reconstruction experiment.
% src(Betti recovery, 9/9 pass): D:\Project\CG-Soup-Topology\tests\test_betti.py
\begin{table}[h]
\centering
\begin{tabular}{lccl}
\toprule
target & $(b_0,b_1,b_2)$ & decisive dim & role \\
\midrule
sphere        & $(1,0,1)$ & H2 & enclosed void \\
cube          & $(1,0,1)$ & H2 & void, flat faces / sharp edges \\
cylinder      & $(1,0,1)$ & H2 & void, capped \\
torus         & $(1,2,1)$ & H1 & loops / handle \\
two\_spheres  & $(2,0,2)$ & H0 & two components \\
three\_spheres& $(3,0,3)$ & H0 & three components \\
double\_torus & $(1,4,1)$ & H1 & genus-2, multi-loop \\
\bottomrule
\end{tabular}
\caption{Synthetic targets and their (verified) Betti numbers.}
% src: D:\Project\CG-Soup-Topology\tests\test_betti.py (sphere/torus/two_spheres/
% double_torus/three_spheres/thick_shell cases) and topology/meshes.py.
% TODO(human): cube/cylinder Betti are verified via the scene gt_mesh.ply target
% diagrams (betti=(1,0,1)); state that provenance if a reviewer asks.
\label{tab:targets}
\end{table}

\paragraph{Metric.}
The primary metric is the \emph{bottleneck
distance}~\cite{cohensteiner2007stability,edelsbrunner2010book} between the reconstruction's
persistence diagram and the target's, in the shape's decisive homology dimension,
reported as the tail mean over the last 30\% of training steps (lower is better).
Geometry is reported in parallel as Chamfer / Hausdorff-95 (\% of bbox diagonal)
so that any topological gain can be checked to occur at \emph{comparable}
geometry rather than from a geometric change.
% src(metric definition): topology/metrics.py (topology_stability_series, bottleneck_distance)

\paragraph{Determinism and seeds.}
Resampling decisions (the face structure $F$) are deterministic given the seed,
but the CUDA rasterizer's gradient atomics make vertex positions $V$
\emph{not} bit-reproducible run-to-run: two same-seed baseline runs can drift by a
small number of triangles. We therefore (i) use \emph{paired} seeds across
conditions and (ii) average each comparison over multiple seeds---5 for the H2
study and the torus, 3 for the remaining crossover shapes; the topology metric
samples the soup $\alpha\!\times$area-weighted and is robust to a few drifting
triangles.
% src(determinism caveat): D:\Project\CG-Soup-Topology\PHASE2_STATUS.md
% src(seed counts, VERIFIED from disk 2026-07-03): h2_unified = 5 seed dirs per
%   shape x condition (all B0-B5); crossover_report records: torus nseeds=5,
%   two_spheres/three_spheres 3, double_torus 3 (exp crossover_N2000).

\paragraph{Budget.}
All comparisons are budget-neutral: the triangle count $N$ is held fixed across
conditions and across the run (the densify step trades faces 1-for-1). Training
runs 2500 steps, at $N{=}1200$ for the H2 study and at each shape's headroom
budget for the crossover (Table~\ref{tab:crossover}).
% src(N=1200, 2500 steps): D:\Project\CG-Soup-Topology\PHASE2_STATUS.md
% src(2500 steps, VERIFIED from disk 2026-07-03): trajectory dumps end at
%   step_02500.pt in h2_unified AND crossover run dirs (torus, two_spheres,
%   double_torus checked) -- no experiment used a different schedule.
