% ============================================================================
%  Topology-Correcting Differentiable Triangle-Soup Reconstruction via
%  Persistent Homology  —  PAPER 2 (DRAFT SKELETON, stage 3e)
%  Build:  cd paper2 && latexmk -pdf main.tex
%
%  GROUNDING CONTRACT (same discipline as paper/):
%   * every numeric result carries a "% src:" comment naming the results file
%     it was read from (PHASE3_PLAN.md appendices A-D and
%     output/synth/topo3/report/results.json). No number without a source.
%   * citations: paper-1 verified set + two web-verified additions
%     (leygonie2022framework, nigmetov2024bigsteps — provenance comments in
%     refs.bib). Never a guessed entry. The allocation study is UNCITED by
%     design: it is reported first-hand in Appendix A (author decision
%     2026-07-09).
%   * author-only spots are marked % TODO(human).
% ============================================================================
\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amsmath,amssymb}
\usepackage{xcolor}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}
\graphicspath{{figures/}}

\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[\,TODO: #1\,]}}}

% TODO(human): venue document class + author block reformat, as with paper 1.
\title{Topology-Correcting Differentiable Triangle-Soup Reconstruction\\
       via Persistent Homology}
% Advisor email verified 2026-07-09 from the official faculty page
% (cp.eng.chula.ac.th/en/about/faculty/pizzanuk). First-author address:
% viritphon.1234@gmail.com CONFIRMED by the author for print (2026-07-09).
\author{Viritphon Chongpermwattanapol\thanks{Corresponding author:
  \texttt{viritphon.1234@gmail.com}} \qquad Pizzanu Kanongchaiyos\\[4pt]
  Department of Computer Engineering, Faculty of Engineering,\\
  Chulalongkorn University, Bangkok, Thailand\\[2pt]
  \texttt{viritphon.1234@gmail.com}\quad\texttt{pizzanu.k@chula.ac.th}}
\date{\today}

\begin{document}
\maketitle

% ----------------------------------------------------------------------------
% ABSTRACT — draft; numbers grounded per the contract.
% TODO(human): voice pass.
% ----------------------------------------------------------------------------
\begin{abstract}
Resampling priors can steer \emph{where} a differentiable triangle-soup
reconstruction spends its triangle budget, but our controlled study of that
channel (reported in Appendix~\ref{app:companion}) found their topological
value is carried largely by spatial \emph{width}, and that no prior shape
repairs
one-dimensional features: concentrated priors manufacture phantom handles,
spread priors never beat a random control. We move topology
from the allocation channel into the \emph{objective}: a differentiable
persistence loss $L=\mathrm{photometric}+\lambda(t)\,L_{\mathrm{topo}}$, where
$L_{\mathrm{topo}}$ is an optimally-matched distance between the live surface's
persistence diagram and a fixed target diagram. Gradients flow through a
pair-frozen backward pass: the matched birth/death simplices are re-expressed
as closed-form circumradii of live sample positions --- exact for the
Gabriel simplices that constitute all significant pairs in practice --- and a
recruitment term restores the gradient that optimal matching provably lacks
when a target feature is missing entirely. A single ratio knob $\rho$ sets the
loss scale by gradient-norm calibration; a curriculum proved unnecessary.
% src: PHASE3_PLAN.md Appendix D / topo3/report/results.json
On budget-starved synthetic scenes the loss is topology-specific where a
norm-matched non-topological control is not: enclosed voids improve
$4.0$--$7.9\times$ and --- the channel's distinguishing result --- torus loops
improve $2.3\times$ with zero phantom handles across seeds, while the control
collapses a loop. The two channels compose: loss plus the best resampling
prior reaches $7.6\times$ below baseline at the best Chamfer. Component
counting (H0) is again matched by the generic control --- a third channel
agreeing that this class needs no topological information. All verdicts
replicate, with no per-shape tuning, on three external genus-known meshes
($2.1$--$10.4\times$ at equal-or-better Chamfer).
% src: topo3/report/results.json (spot/bob/fandisk)
Prescription: correct topology in the loss, allocate with a wide prior, and
combine.
\end{abstract}

\input{sections/intro}
\input{sections/related}
\input{sections/method}
\input{sections/setup}
\input{sections/results}
\input{sections/discussion}
\input{sections/limitations}

% The allocation-channel study (our own prior experiments), reported
% first-hand with numbers. Moves to supplementary at venues with strict
% page limits.
\input{sections/appendix_companion}

\bibliographystyle{unsrtnat}
\bibliography{refs}

\end{document}
