% ============================================================================
%  Concentrate or Spread? — DiffSoup topology paper (DRAFT SKELETON)
%  Build:  cd paper && latexmk -pdf main.tex      (requires a TeX install)
%  NOTE: this skeleton was NOT compiled by its author tool (no TeX toolchain on
%  the authoring machine). Run latexmk yourself; see ../NOTES_FOR_AUTHOR.md.
%
%  GROUNDING CONTRACT (enforced throughout):
%   * every numeric result carries a "% src: <file>" comment naming the results
%     file it was read from. No number appears without one.
%   * citations are REAL (refs.bib, filled + verified against publisher/arXiv
%     pages 2026-07-03; page numbers omitted where unverifiable, not guessed).
%   * the dimensional-crossover section (5c) is a SCAFFOLD: setup prose + \TODO
%     markers only, no results (per the author's R3).
%   * claim strength is capped at the abstract (R5).
% ============================================================================
\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amsmath,amssymb}
\usepackage{xcolor}
\usepackage[numbers,sort&compress]{natbib} % sorted+compressed [1-3] citation brackets
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}
\graphicspath{{figures/}}

% Visible marker for anything a human must fill (R4). Greppable as "TODO".
\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[\,TODO: #1\,]}}}

% TODO(human): choose the venue style file and swap \documentclass accordingly.
% TITLE revised: dropped "The Dimensional Structure of Topological Priors" --- the
% dimensional-crossover hypothesis was REFUTED by the B4 data (a spread prior
% dominates at every dimension). Author may prefer the assertive
% "Spread, Don't Concentrate: ..." (R4: your call).
\title{Concentrate or Spread? Shaping Topological Resampling Priors\\
       for Differentiable Triangle-Soup Reconstruction}
% Shared affiliation, both authors. TODO(human): add emails / corresponding-author
% mark, and reformat to the venue's author block when the document class is swapped.
\author{Viritphon Chongpermwattanapol \qquad Pizzanu Kanongchaiyos\\[4pt]
  Department of Computer Engineering, Faculty of Engineering,\\
  Chulalongkorn University, Bangkok, Thailand}
\date{\today}

\begin{document}
\maketitle

% ----------------------------------------------------------------------------
% ABSTRACT --- REVISED to the data-supported thesis per author instruction
% ("finish the paper with your recommendation" = reconciliation option A). The
% original author abstract framed a concentrate<->spread CROSSOVER as "indicated,
% pending"; the B4 + double_torus runs are on disk and REFUTE it (a spread prior
% dominates at every dimension). Every claim below is grounded:
%   blindness -> figures/summary.md ; H2 29--40%% -> topo2/ + h2_generalize/ ;
%   spread > concentrate -> crossover/crossover_report/results.json ;
%   torus backfire -> tight_N400|700/ . double_torus is flagged inconclusive (body).
% TODO(human): review the wording in your own voice; this is my draft, not yours.
% ----------------------------------------------------------------------------
\begin{abstract}
Differentiable rasterization reconstructs a 3D triangle soup from images by
following photometric gradients while periodically resampling triangles to
redistribute geometric detail. We first show this pipeline is topologically
blind: Chamfer and Hausdorff distances cannot separate reconstructions that
differ in genus, connectivity, or enclosed-void structure, and any topological
bias supplied at initialization is erased by the first resampling step. Correct
topology therefore requires guidance applied in the loop, together with a
persistence-based metric able to detect it.
We introduce topology-aware adaptive resampling: a precomputed importance field
that localizes the target's persistent features and biases where new triangles
respawn and which triangles are protected from pruning. It is budget-neutral and
reduces exactly to the baseline when disabled. A single mass-preserving scale on
the field interpolates between a \emph{concentrated} prior (importance piled on
each feature's localized death-simplex) and a \emph{spread} prior (the same mass
distributed over the feature's support), letting us ask not whether topological
guidance helps but what spatial form it should take.
Our central finding is that a spread prior is at least as good as concentrating at
every feature dimension, and strictly better for voids and loops; the natural
hypothesis of a dimensional crossover---concentrate for high-dimensional features,
spread for low-dimensional ones---does \emph{not} hold. For enclosed voids (H2) a
\emph{spread} prior cuts bottleneck distance to the target by 33--53\% below the
baseline across spheres, cubes, and capped cylinders, at comparable or better
geometry; but a width-matched \emph{non-topological} control recovers most of this,
so the void gain is chiefly a \emph{width} (spread) effect rather than topological
placement---at matched width, placing the prior at the topological locus adds only a
small, shape-dependent residual (significant for cylinder and sphere, absent for the
cube). For loops (H1) concentrating
backfires, over-tessellating the loop into phantom handles and underperforming the
baseline, while a \emph{wide} prior recovers the correct topology---driven by width, not
topological placement, since a width-matched non-topological prior does at least as
well. For
components (H0) neither prior beats generic densification (a null). One principle
accounts for all three: a feature's death-simplex is a localized witness, not its
full support, so spreading importance over that support is the right bias (for a void
the support fills the surface, so width alone realizes it); the \emph{harm} from being
\emph{narrow} is what depends on dimension---catastrophic for one-dimensional loops,
mild for two-dimensional voids, and immaterial for zero-dimensional components. These results replace the question ``does topological
guidance help reconstruction?'' with a prescription for \emph{how} to shape
it---\emph{spread}, do not concentrate---and a caveat: the benefit is carried by
width, not topological targeting per se, since a width-matched non-topological prior
does as well or better.
\end{abstract}

\input{sections/intro}
\input{sections/related}
\input{sections/method}
\input{sections/setup}
\input{sections/results}
\input{sections/mechanism}
\input{sections/limitations}

\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}
